Synthesis of Linear Functional Programs

### 2021-07-23 article/article.tex
Formal system and tweaks

% TODO Verificar nomes dos linear logic theoresm...
{\scriptsize
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Group & Goal & Avg. time $\pm\,\sigma$ & Keywords & Components \\
        \hline
        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & $133\mu s\pm 4.9\mu s$ &&\\ 
        & distributivity & $179\mu s\pm 5.0\mu s$ && \\ 
        & call by name & $196\mu s\pm 4.6\mu s$ && \\ 
        & 0/1 & $294\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{7}{4em}{List} & map & $288\mu s\pm 7.2\mu s$  && \\ 
        & append & $292\mu s\pm 7.0\mu s$ & & \\ 
        & foldl & $1.69ms\pm 5.3\mu s$ & \emph{choose 1} & \\ 
        & foldr & $704\mu s\pm 10\mu s$ && \\ 
        & concat & $505\mu s\pm 18\mu s$ && \\
        & uncons & $215\mu s\pm 15\mu s$ && \\
        & reverse & $17.4ms\pm 515\mu s$ & \emph{reverse [1,2] == [2,1]} & \\ 
        \hline
        \multirow{2}{4em}{Maybe} & $>>=$ & $194\mu s\pm 5.3\mu s$ && \\ 
        & maybe & $161\mu s\pm 4.8\mu s$&& \\
        \hline
        \multirow{7}{4em}{State} & runState & $190\mu s\pm 6.8\mu s$ && \\ 
        & $>>=$ & $979\mu s\pm 23\mu s$ && \\
        & $>>=$ & $\infty$ & \emph{using (runState)} & \\
        & get & $133\mu s\pm 3.8\mu s$ && \\
        & put & $146\mu s\pm 3.4\mu s$ && \\
        & modify & $219\mu s\pm 4.9\mu s$ && \\
        & evalState & $156\mu s\pm 4.0\mu s$ && \\
        % \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & map & 1s & yes & yes \\
        % & append & 1s & yes & appends to rightmost leaf \\
        % & join node & 1s & yes & joins at rightmost leaf \\
        % \hline
        % \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & insertWithKey & 1s & yes & ignores function \\
        % & union & 1s & yes & ignores keys \\
        % & mapAccum & 1s & no & ?? \\
        % & map & 1s & yes & yes \\
        % & mapKeys & 1s & yes & yes \\
        \hline
        \multirow{1}{4em}{Misc} & either & $197\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{2}{4em}{Array} & & & \emph{depth 3} & \emph{freeze, foldl} \\
        & array~\ref{sec:overview} & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
        \hline
    \end{tabular}
  \end{center}
}

@@

\section{Formal System}%
\label{sec:final_system}

@@

>>>>

% TODO Verificar nomes dos linear logic theoresm...
{\scriptsize
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Group & Goal & Avg. time $\pm\,\sigma$ & Keywords & Components \\
        \hline
        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & $133\mu s\pm 4.9\mu s$ &&\\ 
        & distributivity & $179\mu s\pm 5.0\mu s$ && \\ 
        & call by name & $196\mu s\pm 4.6\mu s$ && \\ 
        & 0/1 & $294\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{7}{4em}{List} & map & $288\mu s\pm 7.2\mu s$  && \\ 
        & append & $292\mu s\pm 7.0\mu s$ & & \\ 
        & foldl & $1.69ms\pm 5.3\mu s$ & \emph{choose 1} & \\ 
        & foldr & $704\mu s\pm 10\mu s$ && \\ 
        & concat & $505\mu s\pm 18\mu s$ && \\
        & uncons & $215\mu s\pm 15\mu s$ && \\
        & reverse & $17.4ms\pm 515\mu s$ & \emph{reverse [1,2] == [2,1]} & \\ 
        \hline
        \multirow{2}{4em}{Maybe} & $>>=$ & $194\mu s\pm 5.3\mu s$ && \\ 
        & maybe & $161\mu s\pm 4.8\mu s$&& \\
        \hline
        \multirow{7}{4em}{State} & runState & $190\mu s\pm 6.8\mu s$ && \\ 
        & $>>=$ & $979\mu s\pm 23\mu s$ && \\
        & $>>=$ & $\infty$ & \emph{using (runState)} & \\
        & get & $133\mu s\pm 3.8\mu s$ && \\
        & put & $146\mu s\pm 3.4\mu s$ && \\
        & modify & $219\mu s\pm 4.9\mu s$ && \\
        & evalState & $156\mu s\pm 4.0\mu s$ && \\
        % \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & map & 1s & yes & yes \\
        % & append & 1s & yes & appends to rightmost leaf \\
        % & join node & 1s & yes & joins at rightmost leaf \\
        % \hline
        % \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & insertWithKey & 1s & yes & ignores function \\
        % & union & 1s & yes & ignores keys \\
        % & mapAccum & 1s & no & ?? \\
        % & map & 1s & yes & yes \\
        % & mapKeys & 1s & yes & yes \\
        \hline
        Misc & either & $197\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{2}{4em}{Array} & & & \emph{depth 3} & \emph{freeze, foldl} \\
        & array~\ref{sec:overview} & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
        \hline
        Refinements & add3 & $39ms\pm 1.1ms$ & & + \\
        \hline
    \end{tabular}
  \end{center}
}

@@

\section{Formal System}%
\label{sec:final_system}
{\small
\begin{mathparpagebreakable}
    \infer*[right=($\lolli R$)]
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
    \notin \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x . M : A
    \lolli B \Uparrow}
    \infer*[right=($\with R$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A
    \with B \Uparrow}
    \infer*[right=($\Uparrow$R)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
    \infer*[right=($\tensor L$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
    \and y,z \notin \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}
    \infer*[right=($1 L$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}1 \Uparrow\ \vdash\ \textrm{let}\
    \star =
    x\ \textrm{in}\ M : C}
    \mprset{flushleft}
    \infer*[right=($\oplus L$)]
    {
    \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
    y \notin \Delta' \\
    \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
    z \notin \Delta'' \\
    \Delta' = \Delta''
    }
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \oplus B \Uparrow\ \vdash\
    \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y \rightarrow M\ |\
    \textrm{inr}\ z \rightarrow N : C}
    \infer*[right=($\bang L$)]
    {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}\bang A \Uparrow\ \vdash\
    \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}
    \infer*[right=($\Uparrow$L)]
    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
    \textrm{not left asynchronous}}
    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
    \infer*[right=(decideR)]
    {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
    \infer*[right=(decideL)]
    {\Gamma; \Delta/\Delta' ; A \Downarrow\ \vdash C}
    {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
    \infer*[right=(decideL!)]
    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
  \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash
      M\{(\textrm{snd}\ x)/y\} : C}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
     \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
        T \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
    T \and T \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
    \infer*[right=($\forall L$)]
    {
        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
        \\
        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
        \\
        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
    }
    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
    \infer*[right=($?L$)]
    {\textsc{unify}(?\alpha
    \mapsto C, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
    x{:}?\alpha \Downarrow\ \vdash x : C}
    \infer*[right=($\Downarrow L?$)]
    {\textsc{unify}(?\alpha
    \mapsto A, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
    x{:}A \Downarrow\ \vdash x : ?\alpha}
    \mprset{flushleft}
    \infer*[right=(decideL!)]
    {
    (A, C) \notin \Rho_{L!}
    \\
    \textsc{isExist}(C) \Rightarrow |\{u\ |\
    (f,u)\in\Rho_{L!},\textsc{isPoly}(f),\textsc{isExist}(u)\}| < d_e
    \\
    \Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!}');\Gamma, A; \Delta/\Delta' ; A
    \Downarrow\ \vdash C
    }
    {\Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!});\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
    \infer*[right=(adt-init)]
    {\textsc{unify}(T_{\overline\alpha} \mapsto T_{\overline\beta}, \Theta)}
    {\Theta/\Theta,T_{\overline\alpha} \mapsto T_{\overline\beta},\Rho; \Gamma; \Delta/\Delta'; x{:}T_{\overline\alpha} \Downarrow\ \vdash x :
    T_{\overline\beta}}
    \infer*[right=(refR)]
    { \textsc{getModel}(p) = M }
    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta' \vdash M : \{a : A\ \vert\
    p\}\Uparrow }
    \infer*[right=(refL)]
    { \textsc{sat}(p_{a} \Rightarrow p_{b}) }
    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta'; x{:}(\{a : A\ \vert\
    p_{a}\})\Downarrow\ \vdash x : \{b : A\ \vert\ p_{b}\} }
    \infer*[right=($\bang\Downarrow$L)]
    { \Theta/\Theta';\Rho;\Gamma;\Delta; x{:}A\Downarrow\ \vdash M : C}
    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\bang A\Downarrow\ \vdash M : \bang C}
\end{mathparpagebreakable}
}

@@

\paragraph{Maybe}
input:
\begin{minted}{haskell}
data Maybe a = Nothing | Just a;
data List a = Nil | Cons (a * List a);

synth return :: a -o Maybe a;
synth empty :: Maybe a;
synth bind :: Maybe a -o (a -o Maybe b) -> Maybe b;
synth maybe :: b -> (a -o b) -> Maybe a -o b;
\end{minted}
output:
\begin{minted}{haskell}
return :: forall a . (a -o Maybe a);
return = Just;

empty :: forall a . Maybe a;
empty = Nothing;

bind :: forall a b . (Maybe a -o (!(a -o Maybe b) -o Maybe b));
bind c d = case c of
    Nothing ->
      let !e = d in Nothing
  | Just f -> let !g = d in g f;

maybe :: forall a b . (!b -o (!(a -o b) -o (Maybe a -o b)));
maybe c d e = let !f = c in
  let !g = d in
    case e of
        Nothing -> f
      | Just h -> g h;
\end{minted}
